De novo variants in PLCG1 are associated with hearing impairment, ocular pathology, and cardiac defects

Version 1

Abstract

Abstract

Phospholipase C isozymes (PLCs) hydrolyze phosphatidylinositol 4,5-bisphosphate into inositol 1,4,5-trisphosphate and diacylglycerol, important signaling molecules involved in many cellular processes. PLCG1 encodes the PLCγ1 isozyme that is broadly expressed. Hyperactive somatic mutations of PLCG1 are observed in multiple cancers, but only one germline variant has been reported. Here we describe three unrelated individuals with de novo heterozygous missense variants in PLCG1 (p.Asp1019Gly, p.His380Arg, and p.Asp1165Gly) who exhibit variable phenotypes including hearing loss, ocular pathology and cardiac septal defects. To model these variants in vivo, we generated the analogous variants in the Drosophila ortholog, small wing (sl). We created a null allele slT2A and assessed the expression pattern. sl is broadly expressed, including in wing discs, eye discs, and a subset of neurons and glia. Loss of sl causes wing size reductions, ectopic wing veins and supernumerary photoreceptors. We document that mutant flies exhibit a reduced lifespan and age-dependent locomotor defects. Expressing wild-type sl in slT2A mutant rescues the loss-of-function phenotypes whereas expressing the variants causes lethality. Ubiquitous overexpression of the variants also reduces viability, suggesting that the variants are toxic. Ectopic expression of an established hyperactive PLCG1 variant (p.Asp1165His) in the wing pouch causes severe wing phenotypes, resembling those observed with overexpression of the p.Asp1019Gly or p.Asp1165Gly variants, further arguing that these two are gain-of-function variants. However, the wing phenotypes associated with p.His380Arg overexpression are mild. Our data suggest that the PLCG1 de novo heterozygous missense variants are pathogenic and contribute to the features observed in the probands.

Introduction

The inositol lipid-specific phospholipase C (PLC) isozymes are key signaling proteins that play critical roles in transducing signals from hormones, growth factors, neurotransmitters, and many extracellular stimuli 1–3. The PLCs selectively hydrolyze phosphatidylinositol 4,5-bisphosphate (PIP2) into inositol 1,4,5-trisphosphate (IP3) and diacylglycerol (DAG) 4,5. PIP2 functions as a membrane anchor for numerous proteins and affects membrane dynamics and ion transport 6–8. The two products, IP3 and DAG, are important intracellular second messengers involved in Ca2+ signaling regulation and protein kinase C signaling activation, respectively 9,10. Hence, PLC orchestrates diverse cellular processes and behaviors, including cell growth, differentiation, migration, and cell death 11–14. There are at least thirteen PLC isozymes grouped in 6 classes (β, δ, ε, γ, η, ζ) in mammals with similar enzymatic function, but each PLC has its own spectrum of activators, expression pattern, and subcellular distribution 15–17.

PLCG1 [MIM: 172420] encodes the PLCγ1 isozyme. PLCγ1 is directly activated by receptor tyrosine kinases (RTKs) as well as cytosolic receptors coupled to tyrosine kinases 18. Upon tyrosine phosphorylation, PLCγ1 undergos conformational changes that release its autoinhibition upon which it associates with the plasma membrane to bind and hydrolyze its substrates 19–21. There is a second PLCγ isozyme, PLCγ2, encoded by PLCG2 [MIM: 600220]. Although these two isozymes have similar protein structure and activation mechanism, they are differentially expressed and regulated, and play non-redundant roles 22,23. PLCG2 is mostly expressed in cells of the hematopoietic system and mainly functions in immune response, causing human diseases associated with immune disorders 24–27. However, PLCG1 is ubiquitously expressed and is enriched in the central nervous system (CNS) 28. Plcg1 is essential in mice, and a null allele causes embryonic lethality with developmental defects in the vascular, neuronal, and immune system 29,30. PLCG1 has emerged as a possible driver for cell proliferation, and increased expression level of PLCG1 has been observed in breast cancer, colon cancer, and squamous cell carcinoma 31–34. Moreover, hyperactive somatic mutations of PLCG1 have been observed in angiosarcomas and T cell leukemia/lymphomas 35–37. However, the genotype-phenotype association of germline PLCG1 variants has yet to be explored.

Results

Individuals with de novo heterozygous missense variants in PLCG1 exhibit hearing impairment, ocular pathology, and cardiac defects

Here, we report three unrelated individuals with de novo heterozygous missense variants in PLCG1 (GenBank: NM_002660.3). Proband 1 (c.3056A>G, p.Asp1019Gly) and proband 2 (c.1139A>G, p.His380Arg) were identified through the Undiagnosed Diseases Network (UDN) 45 and proband 3 (c.3494A>G, p.Asp1165Gly) was identified via GeneMatcher 46. The probands range in age from 5 years to 20 years. The phenotypes of the probands partially overlap but show a diverse spectrum. Pertinent shared features include deafness (probands 1 and 3), ophthalmologic abnormalities (probands 1 and 2), and cardiac septal defects (probands 1 and 3). Two of the three individuals show abnormal brain MRI (probands 1 and 2), and one has immune defects (proband 3). Additional variants identified in the probands are discussed in supplemental data. [As per medRxiv policy, the whole and detailed case history for the probands have been removed. To obtain more detailed information, please contact the authors]

The proband-associated missense PLCG1 variants are located in conserved protein domains and are predicted to be deleterious

PLCG1 is predicted to be tolerant to loss-of-function alleles with a pLI score 47 of 0.16, suggesting that loss of one copy of the gene is unlikely to cause haploinsufficiency in humans and consistent with the presence of many protein truncating variants in gnomAD 48. However, the missense constraint Z score 47 of PLCG1 is 3.69, suggesting it is intolerant to missense variants. In addition, the prediction based on the DOMINO algorithm indicates that PLCG1 variants are likely to be dominant 49. The in-silico pathogenicity predictions suggest that these variants are likely to be pathogenic (Table S1) based on MARRVEL 50.

The three variants identified from the probands map to different conserved domains of PLCγ1, and each variant affects an amino acid residue that is conserved from flies to humans (Figure 1A and Figure 1B). The p.Asp1019Gly and p.His380Arg variants map to the catalytic core domains, and the p.Asp1165Gly is in the C-terminal C2 domain. PLCγ1 contains other conserved domains including an N-terminal pleckstrin homology (PH) domain, four EF hand motifs, as well as a PLCγ-specific regulatory array that is composed of a split PH domain (sPH), two Src homology 2 (nSH2 and cSH2) domains and a Src homology 3 (SH3) domain.

The PLCG1 homolog, sl is conserved in Drosophila

(A) Schematic of human PLCG1 and fly Sl protein domains and positions of the variants identified in the probands. Domain prediction is based on annotation from NCBI. Alignment of human PLCG1 and the homologous proteins. The variants are marked with boxes.

(B) All the variants affect conserved amino acids (labeled in red). Isoforms for alignment: Human PLCG1 NP_877963.1; Mouse Plcg1 NP_067255.2; Zebrafish plcg1 NP_919388.1; Fly sl NP_476726.2.

(C) Schematic of fly sl genomic span, transcript, alleles and the 92kb genomic rescue (GR) construct.

Loss-of-function alleles of sl including sl2 (13bp deletion 52), slKO (CRISPR-mediated deletion of the gene span 53), and slT2A (T2A cassette inserted in the first intron, 54) are indicated. The T2A cassette in slT2Ais flanked by FRT sites and can be excised by Flippase to revert loss-of-function phenotypes. GAL4 expression in slT2A under the control by sl endogenous promoter can be used to assess sl gene expression pattern by crossing with a UAS-mCherry.nls reporter line, or be used to model patient variants in vivo by crossing with human PLCG1 cDNAs or corresponding fly sl cDNAs. The primer pair used for real-time PCR is indicated.

small wing (sl) is the fly ortholog of human PLCG1

To obtain information about the nature of the proband-associated variants, we utilize Drosophila to model the variants in vivo. Flies have three genes encoding PLC isozymes (Table S2). Among them, small wing (sl) is predicted to be the ortholog of PLCG1 with a DIOPT (DRSC Integrative Ortholog Prediction Tool) score of 17/18 (DIOPT version 9.0) 51. The encoded proteins share 39% identity and 57% similarity and are composed of similar conserved domains (Figure 1A). sl is also predicted to be the ortholog of PLCG2 with a DIOPT score of 12/18. These data suggest that sl corresponds to two human genes encoding the PLCγ isozymes. We generated transgenic flies carrying the UAS-human PLCG1 cDNAs for both the reference (UAS-PLCG1Reference) and the variants (UAS-PLCG1D1019G, UAS-PLCG1H380R, and UAS-PLCG1D1165G). Given the high level of protein sequence homology and the conservation of the affected amino acids (Figure 1B), we also generated transgenic flies for the reference and analogous variants in the fly sl cDNA, namely UAS-slWTand UAS-slvariants (UAS-slD1041G, UAS-slH384R, and UAS-slD1184G). sl is on the X chromosome, and several alleles of sl have been isolated or generated previously, including sl2, slKO and slT2A (Figure 1C). sl2 carries a 13bp deletion in the third exon that leads to a frameshift and early stop gain 52. sl2 is a strong loss-of-function allele that causes small wing size, ectopic wing veins and extra R7 photoreceptors 52. slKOwas generated by CRISPR-mediated genomic editing that removes nearly the entire gene 53. slT2A allele was generated by inserting an FRT-SA-T2A-GAL4-polyA-FRT cassette as an artificial exon into the first coding intron of sl (Figure 1C) 54,55. The polyA arrests transcription, and slT2Ais a strong loss-of-function allele (Figure S1A). The T2A viral sequence triggers ribosomal skipping and leads to the production of GAL4 proteins 56,57 that are expressed in the proper spatial and temporal pattern of sl. This allows us to assess the expression pattern of sl by driving the expression of a UAS-fluorescent protein 54, or to assess the function of variants by expressing the human UAS-reference/variant cDNAs 42,58–62. In addition, the cassette is flanked by two FRT sites so that it can be excised from the cells that express the gene in the presence of UAS-Flippase and revert the mutant phenotypes (Figure 1C) 54.

We first assessed the expression pattern of sl by driving UAS-mCherry.nls (an mCherry that localizes to nuclei) with slT2A. sl is expressed in the 3rd larval wing discs and eye discs (Figure 2A), consistent with the loss-of-function phenotypes observed in the wings and eyes 52. The expression pattern of sl in the wing discs is not homogenous. Higher expression levels are observed in the anterior compartment and along both the anterior/posterior and dorsal/ventral compartment boundaries (Figure 2A). The hemizygous slT2A/Y male flies and the trans-heterozygous slT2A/sl2 or slT2A/slKO female flies show reduced wing size and ectopic wing veins (Figure 2B and Figure S1B), as well as additional photoreceptors in the eye (Figure 2C and Figure S1C). These phenotypes can be rescued by UAS-Flippase or by introducing a genomic rescue construct (Dp(1;3)DC313 63, Figure 1C) that covers the sl locus (Figure 2B and Figure 2C). These data show that all the observed phenotypes in slT2A mutants can be attributed to the loss of sl.

slT2A is a loss-of-function allele causing phenotypes in wing and eye

(A) sl expression in wing and eye. Expression of UAS-mcherry.nls (red) was driven by slT2Ato label the nuclei of the cells that expressed sl. sl is expressed in the 3rd instar larval wing disc (left) and eye disc (right). Higher magnification image of the wing disc pouch region indicated by dashed rectangle is shown. The posterior/anterior and dorsal/ventral compartment boundaries are indicated by dashed lines in yellow. Scale bars, 100μm

(B) slT2A causes wing size reduction and ectopic veins (arrowhead) in hemizygous mutant male flies.

The wing phenotypes can be rescued by introduction of a genomic rescue (GR) construct or the expression of Flippase. Scale bars, 0.5mm. The quantification of adult wing size is shown in the right panel. Each dot represents the measurement of one adult wing sample. Unpaired t test, ∗∗∗∗p < 0.0001, mean ± SEM.

(C) slT2A causes extra photoreceptors (arrows) in the hemizygous mutant flies. The eye phenotype can be rescued by introduction of a genomic rescue (GR) construct. The photoreceptor rhabdomeres stain positive for phalloidin labeling F-actin. Scale bars, 10μm. The quantification is shown in the right panel. Each dot represents the measurement of one retina sample. Unpaired t test, ∗∗∗∗p < 0.0001, mean ± SEM.

sl is expressed in the fly CNS and loss of sl causes longevity and locomotion defects

Given that human PLCG1 is highly expressed in the central nervous system (CNS) 28 and that the probands present with neurologic phenotypes including hearing or vision deficits (Table 1), we investigated the expression pattern and the cell type specificity of sl in the CNS of flies. sl is expressed in the larval CNS as well as the adult brain, and co-staining with the pan-neuronal marker Elav 64 and glial marker Repo 65 show that sl is expressed in a subset of neurons and glia cells in the CNS (Figure 3A). We therefore assessed the longevity and climbing of slT2A flies. Compared to the wild-type w1118 flies, slT2A/Y hemizygous mutant flies show a shortened lifespan and a progressively reduced climbing ability. These phenotypes can be rescued by expression of the wild-type sl cDNA(slT2A/Y; UAS-slWT) (Figure 3B).

sl is expressed in a subset of neurons and glia in the CNS, and loss of sl causes behavioral defects

(A) Expression pattern of sl in the central nervous system observed by slT2A-driven expression of UAS-mCherry.nls reporter (red). In either larval or adult brain, sl is expressed in a subset of fly neurons and glia, which were labeled by pan-neuronal marker Elav (green, upper panel) and pan-glia marker Repo (green, lower panel). Higher magnification images of the regions indicated by dashed rectangles are shown. Scale bars, 20μm in the magnified images, 50μm in other images.

(B) Loss of sl causes behavioral defects in longevity and locomotion. slT2A hemizygous flies have a shorter lifespan than w1118 control flies. The median life span of slT2A and w1118 flies is 40 days and 62 days respectively. The short lifespan in slT2A flies can be rescued by a UAS transgene that expresses wild-type sl cDNA (slWT). Fly locomotion ability was assessed by climbing assay (see methods). slT2A flies at the age of 7 days show reduced locomotion ability and become almost immotile at the age of 35 days. The reduced locomotion ability in slT2A flies can be fully rescued by slWT. For lifespan assay, Longrank test, ****p<0.0001. For climbing assay, each dots represents one vial containing 18-22 flies for test. Unpaired t test, ****p<0.0001.

Functional assays in flies indicate that the PLCG1 variants are toxic

To assess the impact of the variants, we expressed the sl variant cDNAs in the slT2A/Y hemizygous mutant males(slT2A/Y; UAS-slvariants) and compared their rescue ability with the wild-type sl(slT2A/Y; UAS-slWT). As shown in Figure 4A, the slT2A/Y mutant flies (or the ones expressing a UAS-Empty control construct) have a slightly reduced eclosion rate but expression of the slWT cDNA fully rescues the percentage of eclosing progeny as measured by the Mendelian ratio. In contrast, expressing slH384R causes a severe reduction of the number of eclosing flies with very few escapers (slT2A/Y; UAS-slH384R). Moreover, expression of the slD1041G or slD1184G leads to 100% lethality (Figure 4A). These data clearly indicate that the three variants are toxic.

The variant cDNAs are toxic

(A) Summary of the viability of expressing sl or PLCG1 cDNAs in slT2A mutant flies. Cross strategy: heterozygous slT2A female flies were crossed to male flies carrying UAS-cDNAs or control (UAS-Empty) constructs, or crossed to the y w males as an extra control. The percentages of the hemizygous slT2A male progeny expressing different UAS-cDNA constructs were calculated. The expected Mendelian ratio is 0.25 (indicated by the green line in the graph). The dark gray columns represent human cDNAs, the light gray columns represent fly cDNAs, and the clear columns represent controls (y w or UAS-Empty). Each dot represents one independent replicate. Unpaired t test, *p<0.05, ****p < 0.0001, mean ± SEM.

(B) Summary of the viability of ubiquitous overexpression of sl or PLCG1 cDNAs driven by Tub-GAL4. Expression of wild-type fly sl cDNA shows no impact on viability whereas expression of the sl variants reduces viability. Expressing human PLCG1 cDNAs at 29°C is toxic in flies. Expression of PLCG1Referenceor PLCG1H380R causes semi-lethality, whereas expression of PLCG1D1019G or PLCG1D1165G is 100% lethal at early larval stage.

(C) Summary of the viability of ubiquitous overexpression PLCG1 cDNAs using a strong driver Tub-GAL4 (left panel) or a weaker driver da-GAL4 (right panel) at different temperatures. The expression level rises with higher temperatures and decreases with lower temperatures, and the toxic impact on viability correlates with the expression level. Expression of PLCG1D1019Gor PLCG1D1165G causes more severe toxicity than expression of PLCG1Reference or PLCG1H380R.

Given that the expression of these sl cDNAs is performed in a null mutant background,we opted to assess the phenotypes associated with ectopic expression assays by overexpressing the wild-type or variant sl cDNAs using a strong ubiquitous driver, Tub-GAL4. Overexpression of slWT shows no impact on viability. However, overexpression of slH384R leads to ∼36% lethality, while overexpression of slD1041G or slD1184G results in 100% lethality at the L1-L2 larval stage (Figure 4B). Hence, ubiquitous overexpression assays cause similar but less severe lethal phenotypes than the expression assays driven by slT2A. This difference may arise from the different expression patterns and levels of Tub-GAL4 and slT2A. Importantly, slT2A drives expression in the cells that normally express sl. In summary, the data argue that the sl variants are likely to be gain-of-function or neomorphic alleles and that slD1041G and slD1184G are very strong toxic alleles whereas slH384Ris a milder allele.

To compare the sl and PLCG1 associated phenotypes, we conducted similar assays using human PLCG1 cDNAs. Expression of PLCG1Referencein the slT2A/Y mutant flies (slT2A/Y; UAS-PLCG1Reference) (Figure 4A) reduces viability by 90%, while the ubiquitous expression of PLCG1Referenceunder the control of Tub-GAL4 (Tub-GAL4 > UAS-PLCG1Reference) reduces viability by 63% (Figure 4B). These data argue that the reference human PLCG1 cDNA is toxic in flies. Expression of PLCG1H380R in the slT2A/Y mutant flies (slT2A/Y; UAS-PLCG1H380R) leads to a modest increase in lethality when compared to PLCG1Reference(Figure 4A), whereas Tub-GAL4 > UAS-PLCG1H380R is not worse than PLCG1Reference (Figure 4B). In contrast, expression of PLCG1D1019Gor PLCG1D1165G results in 100% lethality using either of the drivers (Figure 4A and 4B). In summary, expressing the human PLCG1 in flies induces toxicity, and PLCG1H380Rhas only slightly increased toxicity when compared to PLCG1Reference. However, the PLCG1D1019G or PLCG1D1165G variants display a severe gain of toxicity.

Since overexpressing fly slWT does not cause viability issues (Figure 4B), the observed toxicity associated with PLCG1Referencein flies may be due to the elevated expression of the human proteins. This hypothesis can be assessed by ectopic expression assays performed by raising the flies at different temperatures. The GAL4-UAS system is highly temperature-dependent since the promoter in the UAS construct contains an Hsp-70 promoter 66: at 29°C the expression level is much higher than at 25°C, and at 22°C the expression level is significantly lower than at 25°C 67. As shown in Figure 4C, the Tub-GAL4>UAS-PLCG1Reference flies exhibit a high lethality ratio when raised at 29°C (∼63% lethal). This lethality ratio decreased to ∼43% when the flies are raised at 25°C whereas the flies are viable at 22°C. This shows that the toxicity is highly dependent on protein levels. Indeed, the expression levels can be further lowered by using a weak ubiquitous GAL4 driver, da-GAL4. The da-GAL4 > UAS-PLCG1Reference or UAS-PLCG1H380R flies are viable at the three tested temperatures (29°C, 25°C, and 22°C). In contrast, da-GAL4 > UAS-PLCG1D1019Gor UAS-PLCG1D1165G flies are lethal at 29°C, semi-lethal (∼80%-85% lethal) at 25°C, and viable at 22°C (Figure 4C, right panel). The animals that escape lethality at 25°C have smaller pupae (Figure S2) and a reduced adult body size, showing that growth is impeded. In summary, these data consistently confirmed that ubiquitous elevated expression levels of the human PLCG1 is toxic in flies.

To assess other phenotypes, we used nub-GAL4 to mostly drive the expression of UAS-PLCG1 or UAS-sl cDNAs in the wing disc, a well-established model to study growth and differentiation in a tissue that is dispensable for viability 68. Overexpression of either PLCG1Reference or slWT in the wing disc leads to slightly smaller wings with a ∼10% reduction in size when compared to expressing the UAS-Empty control (Figure S3). Hence, both loss and overexpression of sl in the wing lead to a size reduction (Figure 2B and S3). This implies a potential dosage-dependent regulation on wing growth by the PLCγ isozymes, although the underlying mechanism is unknown. Overexpression of the PLCG1H380R or slH384R in the wing results in ∼5% reduction in wing size when compared to the PLCG1Referenceor slWT (Figure S3). However, overexpression of PLCG1D1019Gor PLCG1D1165G results in severe wing phenotypes characterized by notched wing margins, fused/thickened veins and reduced wing sizes (Figure 5A). Notably, overexpressing fly slD1041Gleads to very similar morphological defects as the corresponding human PLCG1D1019Gvariant, indicating that the observed wing phenotypes are indeed due to alterations in PLCG1/sl functions. These data also argue that the human PLCG1 functions in the same pathways as sl. Furthermore, overexpressing the slD1184G leads to pupal lethality (Figure 5A). This suggests that this variant has a more severe impact on development than the other variants and is not inconsistent with the observation that nub-GAL4 drives some expression of UAS-cDNA in the nervous system as well 69.

The PLCG1 p.Asp1019Gly and p.Asp1165Gly variants are likely to be gain-of-function alleles

(A) Wing-specific overexpression of PLCG1D1019G or PLCG1D1165G causes severe wing morphology defects including notched margin (arrows) and fused/thickened veins (arrowheads), fully penetrant (the penetration ratio is indicated). Note that overexpression of the fly slD1041G (corresponding to the human PLCG1D1019G) shows similar phenotypes. Scale bars, 0.5mm.

(B) Wing-specific overexpression of an established hyperactive variant PLCG1D1165H causes wing morphology phenotypes similar to, but more severe than that of PLCG1D1019G or PLCG1D1165G. Scale bars, 0.5mm.

We also assessed the effect of ectopic expression of human PLCG1 on eye development using an eye-specific driver eyeless-GAL4 (ey-GAL4). The expression of PLCG1Reference or PLCG1H380R in fly eyes leads to a mild reduction in eye size, while the expression of PLCG1D1019G or PLCG1D1165Gleads to a severe reduction in eye size (Figure S4). In summary, the eye data are consistent with the wing data, showing that PLCG1D1019G and PLCG1D1165G are more toxic than PLCG1Reference. On the other hand, the toxicity of PLCG1H380R is milder and can be distinguished from that of the PLCG1Reference only in certain contexts.

The p.Asp1019Gly and p.Asp1165Gly variants are gain-of-function variants

Previous studies have identified a very strong gain-of-function somatic PLCG1 variant p.Asp1165His in adult T cell leukemia/lymphoma 20,70,71. This variant has been documented to cause a very dramatic increase in phospholipase activity in vitro 20,71. To characterize the impact of this hyperactive variant in vivo, we generated transgenic flies with this variant and tested PLCG1D1165Hin our tissue-specific expression assays. Overexpression of PLCG1D1165Hin the eye using ey-GAL4 causes lethality at 29°C (Figure S4), arguing that it is highly toxic. As shown in Figure 5B, overexpressing PLCG1D1165H in the wing using the nub-GAL4 driver causes lethality at 29°C. However, these flies survive when they are raised at 25°C, yet the wings show severe morphological defects, including notched wing margins, thickened veins as well as reduced wing size. These phenotypes are similar to, but more severe than, the defects observed in the wings overexpressing PLCG1D1019Gor PLCG1D1165G (Figure 5B). These data provide compelling evidence that these two variants are gain-of-function variants.

Discussion

A recent study reported an individual with a de novo heterozygous gain-of-function germline variant in PLCG1, p.Ser1021Phe 72. The proband exhibited an early-onset and severe immune dysregulation with autoimmune and autoinflammatory symptoms. Tao et al. performed in vitro and ex vivo experiments using cultured cell lines transfected with PLCG1 constructs and peripheral blood mononuclear cells from the proband, respectively. They showed that the p.Ser1021Phe variant led to a 1.5-2 fold increase in intracellular IP3 production compared to controls. However, in our in vivo assays, overexpression of PLCG1S1021Fin wings or eyes does not cause obvious phenotypes when compared to PLCG1Reference (Figure S5), which is clearly distinct from the other gain-of-function variants tested in our assays (p.Asp1019Gly, p.Asp1165Gly and p.Asp1165His) (Figure 5B). It is notable that the three probands reported here are discordant for immune-related phenotypes. Proband 1 with the p.Asp1019Gly variant has no immune dysregulation symptoms. Proband 2 with the p.His380Arg variant has a relapsing steroid responsive inflammatory encephalomyelitis, which is very different from the autoimmune symptoms reported by Tao et al 72. Finally, proband 3 with the p.Asp1165Gly variant presents with a T cell lymphocytopenia with recurrent infections suggesting that the individual is immune compromised. Hence, the immune phenotypes in the individuals reported here are very heterogeneous or absent, and inconsistent with an autoimmune disease. In summary, the immune-related phenotypes and their association with PLCG1 variants will need to be explored in more depth.

Previously, studies based on biochemical assays and protein structures provided insights into how the variants studied here may affect the enzymatic activity of PLCγ1 (the established protein structure of full-length rat Plcg1 is shown in Figure 6A). In its basal state, the PLCγ-specific regulatory array (sPH-nSH2-cSH2-SH3) forms autoinhibitory interfaces with the catalytic domains. Upon activation by the RTKs through binding to nSH2, PLCγ1 is phosphorylated, which in turn induces the dissociation of the inhibitory cSH2 domain from the C2 domain. This triggers conformational rearrangements, allowing the enzyme to associate with the membrane and exposing the catalytic domains to allow hydrolysis of PIP2 19–21,73,74. As shown in Figure 6A, the location of the three variants modeled in this study (Asp1019Gly, His380Arg and Asp1165Gly) are either within the catalytic domains or at the intramolecular interfaces. The p.Asp1019Gly and p.Asp1165Gly variants impact crucial residues involved in autoinhibition. Specifically, the p.Asp1019Gly variant affects a residue located at the apex of the hydrophobic ridge within the Y box (Figure 6B), which is important for the interaction between the sPH domain and the Y box. This interaction is critical for the autoinhibition by blocking the membrane engagement of the catalytic core domain prior to enzymatic activation 20,75. Notably, substitution of Asp1019 with Lys (D1019K) has been demonstrated to enhance basal phospholipase activity by approximately 15 fold in vitro 76. Similarly, another hotspot somatic variant, p.Ser345Phe, located at the hydrophobic ridge within the X box, involved in the interaction between the sPH domain and X box, has also been verified to be hyperactive 37,77. In contrast, the p.Ser1021Phe variant described by Tao et al. 72 lies outside the hydrophobic ridge of the interface between the sPH domain and Y box (Figure 6B). On the other hand, the p.Asp1165Gly variant affects a residue situated within a loop of the C2 domain (Figure 6C). The Asp1165 residue plays a key role in stabilizing the interaction between the cSH2 domain and the C2 domain to maintain the autoinhibited state 78. As mentioned above, the somatic variant p.Asp1165His leads to significantly elevated phospholipase activity in vitro 71,73, and results in severe phenotypes in vivo (Figure 5B). In contrast, the p.His380Arg variant impacts the His380 residue within the X box, situated near the bound Ca2+ cofactor in the catalytic core (Figure 6D). His380 plays a role in coordination of phosphate 1 of IP3 74. While this residue may not be key to the autoinhibition, it is important to the phospholipase activity. Substitution of His380 with Phe or Ala (H380F, H380A) has been reported to suppress PIP2 hydrolysis and IP3 production 79,80. Substitution of the His380 with Arg in p.His380Arg variant may create a more basic environment, impacting the lipase activity in a distinct way. Our in vivo data are consistent with the published in vitro data, strengthening our conclusion that the variants are pathogenic and impact the protein function.

PLCG1 variants affect important residues

(A) 3D structure of full-length rat Plcg1 (PDB code: 6PBC; rat Plcg1 shares 96.9% amino acid identity with human PLCG1). The conserved protein domains are labeled with different colors. Two major intracellular interfaces are circled by dashed lines: 1-The hydrophobic ridge between the sPH domain and the catalytic core (X-box and Y-box); and 2-The interface between the cSH2 domain and the C2 domain. The three residues affected by the variants are indicated by yellow balls.

(B) Enlarged views of the Asp1019 residue within the autoinhibition interface between sPH domain and the Y box. The potential interactions with nearby residues are indicated. Notably, the Ser1021 residue is outside of the hydrophobic ridge and shows no predicted interaction with the residues in the sPH domain.

(C) Enlarged views of the Asp1165 residue within the autoinhibition interface between the cSH2 domain and the C2 domain. The potential interactions with nearby residues are indicated.

(D) Enlarged view of the His380 residue within the X-box catalytic domain, in proximity to the Ca2+ cofactor. Structural analysis was performed via UCSF Chimera 83.

In summary, we report three individuals with de novo heterozygous missense variants in PLCG1 presenting with disease features that encompass ophthalmologic, hearing, and cardiac defects with variable expressivity. Our functional assays provide compelling in vivo evidence that the PLCG1 variants alter normal protein function. However, additional genetic variants (see Table 1) and environmental factors may contribute to some of the diverse phenotypes observed in these three individuals. Moreover, the PLCγ1 isozyme is an integral component of multiple signaling pathways, and the outcomes of its dysregulation are expected to be context dependent. Hence, different variants of PLCG1 may impact diverse cellular processes in different tissues and cells, leading to a range of pathological changes. Even when the affected residues are in close proximity to each other, their impact on protein function can be different 73. Moreover, given that the p.Asp1019Gly and p.Asp1165Gly variants are hyperactive, potential therapeutic targets include specific inhibitors of PLCG1, such as antisense approaches that target the altered nucleotides 81,82.

Data and code availability

This study did not generate datasets. All reagents developed in this study are available upon request.

Conflict of Interest Statement

The Department of Molecular and Human Genetics at Baylor College of Medicine receives revenue from clinical genetic testing completed at Baylor Genetics Laboratories.

Supporting information

eLife Assessment

This important study reveals how Drosophila may be used to investigate the role of missense variants in the gene PLCG1 related to human disease in case studies. The evidence that most of these variants have a gain-of-function effect in the fly is convincing and supportive of their pathogenic effect. With some additional control experiments to assess overexpression toxicity, this work would be of relevance to human and Drosophila geneticists alike.

Reviewer #1 (Public Review):

Summary:

This manuscript provides an initial characterization of three new missense variants of the PLCG1 gene associated with diverse disease phenotypes, utilizing a Drosophila model to investigate their molecular effects in vivo. Through the meticulous creation of genetic tools, the study assesses the small wing (sl) phenotype - the fly's ortholog of PLCG1 - across an array of phenotypes from longevity to behavior in both sl null mutants and variants. The findings indicate that the Drosophila PLCG1 ortholog displays aberrant functions. Notably, it is demonstrated that overexpression of both human and Drosophila PLCG1 variants in fly tissue leads to toxicity, underscoring their pathogenic potential in vivo.

Strengths:

The research effectively highlights the physiological significance of sl in Drosophila. In addition, the study establishes the in vivo toxicity of disease-associated variants of both human PLCG1 and Drosophila sl.

Weaknesses:

The study's limitations include the human PLCG1 transgene's inability to compensate for the Drosophila sl null mutant phenotype, suggesting potential functional divergence between the species. This discrepancy signals the need for additional exploration into the mechanistic nuances of PLCG1 variant pathogenesis, especially regarding their gain-of-function effects in vivo.

Overall:

The study offers compelling evidence for the pathogenicity of newly discovered disease-related PLCG1 variants, manifesting as toxicity in a Drosophila in vivo model, which substantiates the main claim by the authors. Nevertheless, a deeper inquiry into the specific in vivo mechanisms driving the toxicity caused by these variants in Drosophila could significantly enhance the study's impact.

Reviewer #2 (Public Review):

The manuscript by Ma et al. reports the identification of three unrelated people who are heterozygous for de novo missense variants in PLCG1, which encodes phospholipase C-gamma 1, a key signaling protein. These individuals present with partially overlapping phenotypes including hearing loss, ocular pathology, cardiac defects, abnormal brain imaging results, and immune defects. None of the patients present with all of the above phenotypes. PLCG1 has also been implicated as a possible driver for cell proliferation in cancer.

The three missense variants found in the patients result in the following amino acid substitutions: His380Arg, Asp1019Gly, and Asp1165Gly. PLCG1 (and the closely related PLCG2) have a single Drosophila ortholog called small wing (sl). sl-null flies are viable but have small wings with ectopic wing veins and supernumerary photoreceptors in the eye. As all three amino acids affected in the patients are conserved in the fly protein, in this work Ma et al. tested whether they are pathogenic by expressing either reference or patient variant fly or human genes in Drosophila and determining the phenotypes produced by doing so.

Expression in Drosophila of the variant forms of PLCG1 found in these three patients is toxic; highly so for Asp1019Gly and Asp1165Gly, much more modestly for His380Arg. Another variant, Asp1165His which was identified in lymphoma samples and shown by others to be hyperactive, was also found to be toxic in the Drosophila assays. However, a final variant, Ser1021Phe, identified by others in an individual with severe immune dysregulation, produced no phenotype upon expression in flies.

Based on these results, the authors conclude that the PLCG1 variants found in patients are pathogenic, producing gain-of-function phenotypes through hyperactivity. In my view, the data supporting this conclusion are robust, despite the lack of a detectable phenotype with Ser1021Phe, and I have no concerns about the core experiments that comprise the paper.

Figure 6, the last in the paper, provides information about PLCG1 structure and how the different variants would affect it. It shows that the His380, Asp1019, and Asp1165 all lie within catalytic domains or intramolecular interfaces and that variants in the latter two affect residues essential for autoinhibition. It also shows that Ser1021 falls outside the key interface occupied by Asp1019, but more could have been said about the potential effects of Ser1021Phe.

Overall, I believe the authors fully achieved the aims of their study. The work will have a substantial impact because it reports the identification of novel disease-linked genes, and because it further demonstrates the high value of the Drosophila model for finding and understanding gene-disease linkages.

Reviewer #3 (Public Review):

Summary:

The paper attempts to model the functional significance of variants of PLCG2 in a set of patients with variable clinical manifestations.

Strengths:

A study attempting to use the Drosophila system to test the function of variants reported from human patients.

Weaknesses:

Additional experiments are needed to shore up the claims in the paper. These are listed below.

Major Comments:

(1) Does the pLI/ missense constraint Z score prediction algorithm take into consideration whether the gene exhibits monoallelic or biallelic expression?

(2) Figure 1B: Include human PLCG2 in the alignment that displays the species-wide conserved variant residues.

(3) Figure 4A:

Given that

(i) sl is predicted to be the fly ortholog for both mammalian PLCγ isozymes: PLCG1 and PLCG2 [Line 62]

(ii) they are shown to have non-redundant roles in mammals [Line 71] and

(iii) reconstituting PLCG1 is highly toxic in flies, leading to increased lethality.

This raises questions about whether sl mutant phenotypes are specifically caused by the absence of PLG1 or PLCG2 functions in flies. Can hPLCG2 reconstitution in sl mutants be used as a negative control to rule out the possibility of the same?

(4) Do slT2A/Y; UAS-PLCG1Reference flies survive when grown at 22{degree sign}C? Since transgenic fly expressing PLCG1 cDNA when driven under ubiquitous gal4s, Tubulin and Da, can result in viable progeny at 22{degree sign}C, the survival of slT2A/Y; UAS-PLCG1Reference should be possible.

and similarly

Does slT2A flies exhibit the phenotypes of (i) reduced eclosion rate (ii) reduced wing size and ectopic wing veins and (iii) extra R7 photoreceptor in the fly eye at 22{degree sign}C?

If so, will it be possible to get a complete rescue of the slT2A mutant phenotypes with the hPLCG1 cDNA at 22{degree sign}C? This dataset is essential to establish Drosophila as an ideal model to study the PLCG1 de novo variants.

(5) Localisation and western blot assays to check if the introduction of the de novo mutations can have an impact on the sub-cellular targeting of the protein or protein stability respectively.

(6) Analysing the nature of the reported gain of function (experimental proof for the same is missing in the manuscript) variants:

Instead of directly showing the effect of introducing the de novo variant transgenes in the Drosophila model especially when the full-length PLCG1 is not able to completely rescue the slT2A phenotype;

(i) Show that the gain-of-function variants can have an impact on the protein function or signalling via one of the three signalling outputs in the mammalian cell culture system: (i) inositol-1,4,5-trisphosphate production, (ii) intracellular Ca2+ release or (iii) increased phosphorylation of extracellular signal-related kinase, p65, and p38.

OR

(ii) Run a molecular simulation to demonstrate how the protein's auto-inhibited state can be disrupted and basal lipase activity increased by introducing D1019G and D1165G, which destabilise the association between the C2 and cSH2 domains. The H380R variant may also exhibit characteristics similar to the previously documented H335A mutation which leaves the protein catalytically inactive as the residue is important to coordinate the incoming water molecule required for PIP2 hydrolysis.

(7) Clarify the reason for carrying out the wing-specific and eye-specific experiments using nub-gal4 and eyless-gal4 at 29˚C despite the high gal4 toxicity at this temperature.

(8) For the sake of completeness the authors should also report other variants identified in the genomes of these patients that could also contribute to the clinical features.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

This manuscript provides an initial characterization of three new missense variants of the PLCG1 gene associated with diverse disease phenotypes, utilizing a Drosophila model to investigate their molecular effects in vivo. Through the meticulous creation of genetic tools, the study assesses the small wing (sl) phenotype - the fly's ortholog of PLCG1 - across an array of phenotypes from longevity to behavior in both sl null mutants and variants. The findings indicate that the Drosophila PLCG1 ortholog displays aberrant functions. Notably, it is demonstrated that overexpression of both human and Drosophila PLCG1 variants in fly tissue leads to toxicity, underscoring their pathogenic potential in vivo.

Strengths:

The research effectively highlights the physiological significance of sl in Drosophila. In addition, the study establishes the in vivo toxicity of disease-associated variants of both human PLCG1 and Drosophila sl.

Weaknesses:

The study's limitations include the human PLCG1 transgene's inability to compensate for the Drosophila sl null mutant phenotype, suggesting potential functional divergence between the species. This discrepancy signals the need for additional exploration into the mechanistic nuances of PLCG1 variant pathogenesis, especially regarding their gain-of-function effects in vivo.

Overall:

The study offers compelling evidence for the pathogenicity of newly discovered disease-related PLCG1 variants, manifesting as toxicity in a Drosophila in vivo model, which substantiates the main claim by the authors. Nevertheless, a deeper inquiry into the specific in vivo mechanisms driving the toxicity caused by these variants in Drosophila could significantly enhance the study's impact.

Reviewer #2 (Public Review):

The manuscript by Ma et al. reports the identification of three unrelated people who are heterozygous for de novo missense variants in PLCG1, which encodes phospholipase C-gamma 1, a key signaling protein. These individuals present with partially overlapping phenotypes including hearing loss, ocular pathology, cardiac defects, abnormal brain imaging results, and immune defects. None of the patients present with all of the above phenotypes. PLCG1 has also been implicated as a possible driver for cell proliferation in cancer.

The three missense variants found in the patients result in the following amino acid substitutions: His380Arg, Asp1019Gly, and Asp1165Gly. PLCG1 (and the closely related PLCG2) have a single Drosophila ortholog called small wing (sl). sl-null flies are viable but have small wings with ectopic wing veins and supernumerary photoreceptors in the eye. As all three amino acids affected in the patients are conserved in the fly protein, in this work Ma et al. tested whether they are pathogenic by expressing either reference or patient variant fly or human genes in Drosophila and determining the phenotypes produced by doing so.

Expression in Drosophila of the variant forms of PLCG1 found in these three patients is toxic; highly so for Asp1019Gly and Asp1165Gly, much more modestly for His380Arg. Another variant, Asp1165His which was identified in lymphoma samples and shown by others to be hyperactive, was also found to be toxic in the Drosophila assays. However, a final variant, Ser1021Phe, identified by others in an individual with severe immune dysregulation, produced no phenotype upon expression in flies.

Based on these results, the authors conclude that the PLCG1 variants found in patients are pathogenic, producing gain-of-function phenotypes through hyperactivity. In my view, the data supporting this conclusion are robust, despite the lack of a detectable phenotype with Ser1021Phe, and I have no concerns about the core experiments that comprise the paper.

Figure 6, the last in the paper, provides information about PLCG1 structure and how the different variants would affect it. It shows that the His380, Asp1019, and Asp1165 all lie within catalytic domains or intramolecular interfaces and that variants in the latter two affect residues essential for autoinhibition. It also shows that Ser1021 falls outside the key interface occupied by Asp1019, but more could have been said about the potential effects of Ser1021Phe.

Overall, I believe the authors fully achieved the aims of their study. The work will have a substantial impact because it reports the identification of novel disease-linked genes, and because it further demonstrates the high value of the Drosophila model for finding and understanding gene-disease linkages.

Reviewer #3 (Public Review):

Summary:

The paper attempts to model the functional significance of variants of PLCG2 in a set of patients with variable clinical manifestations.

Strengths:

A study attempting to use the Drosophila system to test the function of variants reported from human patients.

Weaknesses:

Additional experiments are needed to shore up the claims in the paper. These are listed below.

Major Comments:

(1) Does the pLI/ missense constraint Z score prediction algorithm take into consideration whether the gene exhibits monoallelic or biallelic expression?

To our knowledge, pLI and missense Z don't consider monoallelic or biallelic expression. Instead, they reflect sequence constraint and are calculated based on the observed versus expected variant frequencies in population databases.

(2) Figure 1B: Include human PLCG2 in the alignment that displays the species-wide conserved variant residues.

We have updated Figure 1B and incorporated the alignment of PLCG2.

(3) Figure 4A:

Given that

(i) sl is predicted to be the fly ortholog for both mammalian PLCγ isozymes: PLCG1 and PLCG2 [Line 62]

(ii) they are shown to have non-redundant roles in mammals [Line 71]

(iii) reconstituting PLCG1 is highly toxic in flies, leading to increased lethality.

This raises questions about whether sl mutant phenotypes are specifically caused by the absence of PLCG1 or PLCG2 functions in flies. Can hPLCG2 reconstitution in sl mutants be used as a negative control to rule out the possibility of the same?

The studies about the non-redundant roles of PLCG1 and PLCG2 mainly concern the immune system.

We have assessed the phenotypes in the slT2A/Y; UAS-hPLCG2 flies. Expression of human PLCG2 in flies is also toxic and leads to severely reduced eclosion rate.

We have updated the manuscript with these results, and included the eclosion rate of slT2A/Y; UAS-hPLCG2 flies in the new Figure 4B.

(4) Do slT2A/Y; UAS-PLCG1Reference flies survive when grown at 22{degree sign}C? Since transgenic fly expressing PLCG1 cDNA when driven under ubiquitous gal4s, Tubulin and Da, can result in viable progeny at 22{degree sign}C, the survival of slT2A/Y; UAS-PLCG1Reference should be possible.

The eclosion rate of slT2A/Y >PLCG1Reference flies at 22°C is slightly higher than at 25°C, but remains severely reduced compared to the UAS-Empty control. We have presented these results in the updated Figure S3.

and similarly

Does slT2A flies exhibit the phenotypes of (i) reduced eclosion rate (ii) reduced wing size and ectopic wing veins and (iii) extra R7 photoreceptor in the fly eye at 22{degree sign}C?

The mutant phenotypes are still observed at 22 °C.

If so, will it be possible to get a complete rescue of the slT2A mutant phenotypes with the hPLCG1 cDNA at 22{degree sign}C? This dataset is essential to establish Drosophila as an ideal model to study the PLCG1 de novo variants.

Thank you for the suggestion. It is difficult to directly assess the rescue ability of the PLCG1 cDNAs due to the toxicity. However, our ectopic expression assays show that the variants are more toxic than the reference with variable severities, suggesting that the variants are deleterious.

The ectopic expression strategy has been used to evaluate the consequence of genetic variants and has significantly contributed to the interpretation of their pathogenicity in many cases (reviewed in Her et al., Genome, 2024, PMID: 38412472).

(5) Localisation and western blot assays to check if the introduction of the de novo mutations can have an impact on the sub-cellular targeting of the protein or protein stability respectively.

Thank you for the suggestion.

We expressed PLCG1 cDNAs in the larval salivary glands and performed antibody staining (rabbit anti-Human PLCG1; 1:100, Cell Signaling Technology, #5690). The larval salivary gland are composed of large columnar epithelia cells that are ideal for analyzing subcellular localization of proteins. The PLCG1 proteins are cytoplasmic and localize near the cell surface, with some enrichment in the plasma membrane region. The variant proteins are detected, and did not show significant difference in expression level or subcellular distribution compared to the reference. We did not include this data.

(6) Analysing the nature of the reported gain of function (experimental proof for the same is missing in the manuscript) variants:

Instead of directly showing the effect of introducing the de novo variant transgenes in the Drosophila model especially when the full-length PLCG1 is not able to completely rescue the slT2A phenotype;

(i) Show that the gain-of-function variants can have an impact on the protein function or signalling via one of the three signalling outputs in the mammalian cell culture system: (i) inositol-1,4,5-trisphosphate production, (ii) intracellular Ca2+ release or (iii) increased phosphorylation of extracellular signal-related kinase, p65, and p38.

We appreciate the reviewer’s suggestion. We utilized the CaLexA (calcium-dependent nuclear import of LexA) system (Masuyama et al., J Neurogenet, 2012, PMID: 22236090) to assess the intracellular Ca2+ change associated with the expression of PLCG1 cDNAs in fly wing discs. The results show that, compared to the reference, expression of the D1019G or D1165G variants leads to elevated intracellular Ca2+ levels, similar to the hyperactive S1021F and D1165H variants. However, the H380R or L597F variants did not show a detectable phenotype in this assay. These results suggest that D1019G and D1165G are hyperactive variants, whereas H380R and L597F variant are not, or their effect is too mild to be detected in this assay. We have updated the related sections in the manuscript and Figures 5A and S5.

OR

(ii) Run a molecular simulation to demonstrate how the protein's auto-inhibited state can be disrupted and basal lipase activity increased by introducing D1019G and D1165G, which destabilise the association between the C2 and cSH2 domains. The H380R variant may also exhibit characteristics similar to the previously documented H335A mutation which leaves the protein catalytically inactive as the residue is important to coordinate the incoming water molecule required for PIP2 hydrolysis.

We utilized the DDMut platform, which predicts changes in the Gibbs Free Energy (ΔΔG) upon single and multiple point mutations (Zhou et al., Nucleic Acid Res, 2023, PMID: 37283042), to gain insight into the molecular dynamics changes of variants. The results are now presented in Figure S7.

Additionally, we performed Molecular dynamics (MD) simulations. The results show that, similar to the hyperactive D1165H variant, the D1019G and D11656G variants exhibit increased disorganization, with a higher root mean square deviations (RMSD) compared to the reference PLCG1.The data are also presented in the updated Figure S7.

(7) Clarify the reason for carrying out the wing-specific and eye-specific experiments using nub-gal4 and eyless-gal4 at 29˚C despite the high gal4 toxicity at this temperature.

We used high temperature and high expression level to see if the mild H380R and L597F variants could show phenotypes in this condition.

The toxicity of the two strong variants (D1019G and D1165G) has been consistently confirmed in multiple assays at different temperatures.

(8) For the sake of completeness the authors should also report other variants identified in the genomes of these patients that could also contribute to the clinical features.

Thank you!

The additional variants and their potential contributions to the clinical features are listed and discussed in Table 1 and its legend.

Reviewer #1 (Recommendations For The Authors):

The manuscript's significant contribution is tempered by a lack of comprehensive analysis using the generated genetic reagents in Drosophila. To enhance our understanding of the PLCG1 orthologs, I suggest the following:

(1) A more detailed molecular analysis to distinguish the actions of sl variants from the wild-type could be very informative. For example, utilizing the HA-epitope tag within the current UAS-transgenes could reveal more about the cellular dynamics and abundance of these variants, potentially elucidating mechanisms beyond gain-of-function.

We appreciate the reviewer’s suggestion. The UAS-sl cDNA constructs contain stop codon and do not express an HA-epitope tag. Alternatively, we utilized commercially available antibodies against human PLCG1 antibodies to assess the subcellular localization and protein stability by expressing the reference and variant PLCG1 cDNAs in Drosophila larval salivary glands. The reference proteins are cytoplasmic with some enrichment along the plasma membrane. However, we did not observe significant differences between the reference and variant proteins in this assay. We did not include this data.

(2) I suggest further investigating the relative contributions of developmental processes and acute (Adult) effects on the sl-variant phenotypes observed. For example, employing systems that allow for precise temporal control of gene expression, such as the temperature-sensitive Gal80, could differentiate between these effects, shedding light on the mechanisms that affect longevity and locomotion. This knowledge would be vital for a deeper understanding of the corresponding human disorders and for developing therapeutic interventions.

We appreciate the reviewer’s suggestion. We utilized Tub-GAL4, Tub-GAL80ts to drive the expression of sl wild-type or variant cDNAs, and performed temperature shifts after eclosion to induce expression of the cDNAs only in adult flies. The slD1184G variant (corresponding to PLCG1D1165G) caused severely reduced lifespan and the flies mostly die within 10 days. The slD1041G variant (corresponding to PLCG1D1019G) led to reduced longevity and locomotion. The slH384R variant (corresponding to PLCG1H380R) showed only a mild effect on longevity and no significant effect on climbing ability. These results suggest that the two strong variants (slD1041G and slD1184G) contribute to both developmental and acute effects while the H384R variant mainly contributes to developmental stages.

I also suggest a more refined analysis of overexpression toxicity. Rather than solely focusing on ubiquitous transgene expression, overexpressing transgene in endogenous pattern using sl-t2a-Gal4 may yield a more nuanced understanding of the pathogenic mechanisms of gain-of-function mutations, particularly in the pathogenesis associated with these variants exclusively located in the coding regions.

We appreciate the reviewer’s suggestion. We therefore performed the experiments using slT2A to drive overexpression ofPLCG1cDNAs in heterozygous female progeny with one copy of wild-type sl+ (slT2A/ yw > UAS-cDNAs). In this context, expression of PLCG1Reference, PLCG1H380RorPLCG1L597F is viable whereas expression of PLCG1D1019G or PLCG1D1165G is lethal, suggesting that the PLCG1D1019G and PLCG1D1165G variants exert a strong dominant toxic effect while the PLCG1H380Rand PLCG1L597F are comparatively milder. Similar patterns have been consistently observed in other ectopic expression assays with varying degrees of severity. These results are updated in the manuscript and figures.

Reviewer #2 (Recommendations For The Authors):

The work in the paper could be usefully extended by determining the effects of expressing His380Phe and His380Ala in flies. These variants suppress PLCG1 activity, so their phenotype, if any, would be predicted not to be the same as His380Arg. Determining this would add further strength to the conclusions of the paper.

We thank the reviewer for the constructive suggestions! We have tested the enzymatic-dead H380A variant, which still exhibits toxicity when expressed in slT2A/Y hemizygous flies, but it is not toxic in heterozygous females suggesting that the reduced eclosion rate is likely not directly associated with enzymatic activity. We have updated the manuscript and figures accordingly.

Version 2

Abstract

Abstract

Phospholipase C isozymes (PLCs) hydrolyze phosphatidylinositol 4,5-bisphosphate (PIP2) into inositol 1,4,5-trisphosphate (IP3) and diacylglycerol (DAG), important signaling molecules involved in many cellular processes including Ca2+ release from the endoplasmic reticulum (ER). PLCG1 encodes the PLCγ1 isozyme that is broadly expressed. Hyperactive somatic mutations of PLCG1 are observed in multiple cancers, but only one germline variant has been reported. Here we describe seven individuals with heterozygous missense variants in PLCG1 [p.(Asp1019Gly), p.(His380Arg), p.(Asp1165Gly), and p.(Leu597Phe)] who present with hearing impairment (5/7), ocular pathology (4/7), cardiac septal defects (3/6), and various immunological issues (5/7). To model these variants in vivo, we generated the analogous variants in the Drosophila ortholog, small wing (sl). We created a null allele slT2A and assessed its expression pattern. sl is broadly expressed, including wing discs, eye discs, and a subset of neurons and glia. slT2Amutant flies exhibit wing size reductions, ectopic wing veins, and supernumerary photoreceptors. We document that mutant flies also exhibit a reduced lifespan and age-dependent locomotor defects. Expressing wild-type sl in slT2A mutant flies rescues the loss-of-function phenotypes, whereas the variants increase lethality. Ectopic expression of an established hyperactive PLCG1 variant, p.(Asp1165His) in the wing pouch causes elevated Ca2+ activity and severe wing phenotypes. These phenotypes are also observed when the p.(Asp1019Gly) or p.(Asp1165Gly) variants are overexpressed in the wing pouch, arguing that these are gain-of-function variants. However, the wing phenotypes associated with p.(His380Arg) or p.(Leu597Phe) overexpression are either mild or only partially penetrant. Our data suggest that the heterozygous missense variants reported here affect protein function differentially and contribute to the clinical features observed in the affected individuals.

Introduction

The inositol lipid-specific phospholipase C (PLC) isozymes are key signaling proteins that play critical roles in transducing signals from hormones, growth factors, neurotransmitters, and many extracellular stimuli (Berridge and Irvine 1984; Exton 1996; Balla 2013). The PLCs selectively hydrolyze phosphatidylinositol 4,5-bisphosphate (PIP2) into inositol 1,4,5-trisphosphate (IP3) and diacylglycerol (DAG) (Nishizuka 1984; Majerus et al. 1986). PIP2 functions as a membrane anchor for numerous proteins and affects membrane dynamics and ion transport (Hilgemann et al. 2001; Hilgemann 2007; Suh and Hille 2008). The two products, IP3 and DAG, are important intracellular second messengers involved in Ca2+ signaling regulation and protein kinase C signaling activation, respectively (Nishizuka 1992; Berridge 1993). Hence, PLC orchestrates diverse cellular processes and behaviors, including cell growth, differentiation, migration, and cell death (Yang et al. 2012; Cocco et al. 2015; Gomes et al. 2021; Asano et al. 2022). There are at least thirteen PLC isozymes grouped in 6 classes (β, δ, ε, γ, η, ζ) in mammals with similar enzymatic function, but each PLC has its own spectrum of activators, expression pattern, and subcellular distribution (Suh et al. 2008; Kadamur and Ross 2013; Katan and Cockcroft 2020).

PLCG1 [MIM: 172420] encodes the PLCγ1 isozyme. PLCγ1 can be directly activated by receptor tyrosine kinases (RTKs) as well as cytosolic receptors coupled to tyrosine kinases (Gresset et al. 2012). Upon tyrosine phosphorylation, PLCγ1 undergoes conformational changes that release its autoinhibition upon which it associates with the plasma membrane to bind and hydrolyze its substrates (Gresset et al. 2010; Hajicek et al. 2019; Nosbisch et al. 2022). There is a second PLCγ isozyme, PLCγ2, encoded by PLCG2 [MIM: 600220]. Although these two isozymes have similar protein structure and activation mechanism, they are differentially expressed and regulated, and play non-redundant roles (Homma et al. 1989; Regunathan et al. 2006). PLCG2 is mostly expressed in cells of the hematopoietic system and mainly functions in immune response, causing human diseases associated with immune disorders (Yu et al. 2005; Ombrello et al. 2012; Zhou et al. 2012; Neves et al. 2018; Baysac et al. 2024). However, PLCG1 is ubiquitously expressed and is enriched in the central nervous system (CNS) (Consortium 2015). Plcg1 is essential in mice, and a null allele causes embryonic lethality with developmental defects in the vascular, neuronal, and immune system (Ji et al. 1997; Liao et al. 2002). PLCG1 has emerged as a possible driver for cell proliferation, and increased expression levels of PLCG1 have been observed in breast cancer, colon cancer, and squamous cell carcinoma (Arteaga et al. 1991; Noh et al. 1994; Park et al. 1994; Xie et al. 2010). Moreover, hyperactive somatic mutations of PLCG1 have been observed in angiosarcomas and T cell leukemia/lymphomas (Behjati et al. 2014; Kunze et al. 2014; Vaque et al. 2014; Kataoka et al. 2015). However, the genotype-phenotype association of germline PLCG1 variants has yet to be explored.

Here, we reported seven individuals carrying heterozygous variants in PLCG1 (GenBank: NM_002660.3) who exhibit partially overlapping clinical features including hearing impairment (5/7), ocular pathology (4/7), cardiac defects (3/6), abnormal brain MRI findings (2/3), and immunological issues with diverse manifestations (5/7). Utilizing Drosophila to model the variants in vivo, we provide evidence that the missense PLCG1 variants are toxic and affect protein function to varying degrees. We argue that these variants contribute to the clinical symptoms observed in the affected individuals.

Results

Individuals with heterozygous missense variants in PLCG1 exhibit hearing impairment, cardiac defects, ocular pathology, and immune dysregulation

Seven individuals with heterozygous missense variants in PLCG1 were recruited trough the Undiagnosed Diseases Network (UDN) (Splinter et al. 2018) (Individuals 1-2) and GeneMatcher (Sobreira et al. 2015) (Individuals 3-7). Individuals 1-3 are de novo cases from unrelated families: Individual 1 [c.3056A>G, p.(Asp1019Gly)], Individual 2 [c.1139A>G, p.(His380Arg)] and Individual 3 [c.3494A>G, p.(Asp1165Gly)]. Individuals 4-7 are from the same family, and all carry the PLCG1 variant [c.1789C>T p.(Leu597Phe)]. The phenotypes of the individuals partially overlap but show a spectrum of clinical manifestations. [As per medRxiv policy, the whole and detailed case history for the probands have been removed. To obtain more detailed information, please contact the authors]

The missense PLCG1 variants affect conserved protein domains and are predicted to be deleterious

PLCG1 is predicted to be tolerant to loss-of-function alleles with a pLI score (Lek et al. 2016) of 0.16, suggesting that loss of one copy of the gene is unlikely to cause haploinsufficiency in humans, consistent with the presence of many protein truncating variants in gnomAD (Karczewski et al. 2020). However, the missense constraint Z score (Lek et al. 2016) of PLCG1 is 3.69, suggesting that it is intolerant to missense variants. Consistently, all variants are located within regions or stretches depleted in missense variants according to scores such as regional missense constraint (RMC) (Chao et al. 2024) or missense tolerance ratio (MTR) (Sun et al. 2024). In addition, the prediction based on the DOMINO algorithm indicates that PLCG1 variants are likely to have a dominant effect (Quinodoz et al. 2017). Several other in-silico pathogenicity predictions also suggest that these variants are likely to be pathogenic (Table S1) based on MARRVEL (Wang et al. 2017).

The four variants identified from the affected individuals map to different conserved domains of PLCγ1, and each variant affects an amino acid residue that is conserved from flies to humans (Figure 1A and 1B). The p.(Asp1019Gly) and p.(His380Arg) variants map to the catalytic core domains (X and Y regions, respectively), the p.(Asp1165Gly) variant is in the C-terminal C2 domain and the p.(Leu597Phe) variant is in the nSH2 domain. The latter is part of the PLCγ-specific regulatory array composed of a split PH domain (sPH), two Src homology 2 (nSH2 and cSH2) domains and a Src homology 3 (SH3) domain. PLCγ1 also contains other conserved domains including an N-terminal pleckstrin homology (PH) domain and four EF hand motifs.

The PLCG1 ortholog is small wing (sl) in Drosophila

(A) Schematic of human PLCG1 and fly Sl protein domains and positions of the variants identified in the affected individuals. Domain prediction is based on annotation from NCBI.

(B) Alignment of protein domains near variants of PLCG1 and PLCG2 with PLCG1 from other species. The variants are marked with boxes. All the variants affect conserved amino acids (labeled in red). Isoforms for alignment: Human PLCG1 NP_877963.1; Human PLCG2 NP_002652.2; Mouse Plcg1 NP_067255.2; Zebrafish plcg1 NP_919388.1; Fly sl NP_476726.2.

(C) Schematic of fly sl genomic span, transcript, alleles and the 92kb genomic rescue (GR) construct. Loss-of-function alleles of sl including sl2 (13bp deletion (Thackeray et al. 1998)), slKO (CRISPR-mediated deletion of the gene span (Trivedi et al. 2020)), and slT2A(T2A cassette inserted in the first intron, (Lee et al. 2018)) are indicated. The T2A cassette in slT2A is flanked by FRT sites and can be excised by Flippase to revert loss-of-function phenotypes. GAL4 expression in slT2A is driven by the endogenous sl promoter, allowing assessment of sl gene expression pattern with a UAS-mCherry.nls reporter line. This system also allows in vivo modeling of proband-associated variants by crossing with human PLCG1 cDNAs or corresponding fly sl cDNAs. The primer pair used for real-time PCR is indicated.

The small wing (sl) is the fly ortholog of human PLCG1

Flies have three genes encoding PLC isozymes (Table S2). Among them, small wing (sl) is predicted to be the ortholog of PLCG1 with a DIOPT (DRSC Integrative Ortholog Prediction Tool) score of 17/18 (DIOPT version 9.0) (Hu et al. 2021). The encoded proteins share 39% identity and 57% similarity and are composed of similar conserved domains (Figure 1A). The sl gene is also predicted to be the ortholog of PLCG2 with a DIOPT score of 12/18. These data suggest that sl corresponds to two human genes encoding the PLCγ isozymes. To obtain information about the nature of the PLCG1 variants, we utilize Drosophila to model them in vivo using the binary GAL4 system (Brand and Perrimon 1993). We generated transgenic flies carrying the UAS-human PLCG1 cDNAs for both the reference (UAS-PLCG1Reference) and the variants (UAS-PLCG1D1019G, UAS-PLCG1H380R, UAS-PLCG1D1165G, and UAS-PLCG1L597F). Given the high level of protein sequence homology and the conservation of the affected amino acids (Figure 1B), we also generated transgenic flies for the reference and analogous variants in the fly sl cDNA, namely UAS-slWT and UAS-slvariants (UAS-slD1041G, UAS-slH384R, UAS-slD1184G, and UAS-slL630F).

In Drosophila, sl is on the X chromosome, and several alleles of sl have been isolated or previously generated, including sl2, slKO and slT2A(Figure 1C). sl2 carries a 13bp deletion in the third exon that leads to a frameshift and early stop gain (Thackeray et al. 1998). sl2 is a strong loss-of-function allele that causes small wing size, ectopic wing veins and extra R7 photoreceptors (Thackeray et al. 1998). slKO was generated by CRISPR-mediated genomic editing that removes nearly the entire gene (Trivedi et al. 2020). slT2A allele was generated by inserting an FRT-SA-T2A-GAL4-polyA-FRT cassette as an artificial exon into the first coding intron of sl (Figure 1C) (Diao et al. 2015; Lee et al. 2018). The polyA arrests transcription, and slT2A is a strong loss-of-function allele (Figure S1A). The T2A viral sequence triggers ribosomal skipping and leads to the production of GAL4 proteins (Donnelly et al. 2001; Diao and White 2012) that are expressed in the proper spatial and temporal pattern of sl. This allows us to assess the expression pattern of sl by driving the expression of a UAS-fluorescent protein (Lee et al. 2018), or to assess the function of variants by expressing the human UAS-reference/variant cDNAs (Huang et al. 2022a; Huang et al. 2022b; Lu et al. 2022a; Lu et al. 2022b; Ma et al. 2023; Pan et al. 2023). In addition, the cassette is flanked by two FRT sites and can therefore be excised from the cells that express the gene in the presence of UAS-Flippase to revert the mutant phenotypes (Figure 1C) (Lee et al. 2018).

We first assessed the expression pattern of sl by driving UAS-mCherry.nls (an mCherry that localizes to nuclei) with slT2A. sl is expressed in the 3rd larval wing discs and eye discs (Figure 2A), consistent with the loss-of-function phenotypes observed in the wings and eyes (Thackeray et al. 1998). The expression pattern of sl in the wing discs is not homogenous. Higher expression levels are observed in the anterior compartment and along both the anterior/posterior and dorsal/ventral compartment boundaries (Figure 2A). The hemizygous slT2A/Y male flies and the trans-heterozygous slT2A/sl2 or slT2A/slKOfemale flies show reduced wing size and ectopic wing veins (Figure 2B and Figure S1B), as well as additional photoreceptors in the eye (Figure 2C and Figure S1C). These phenotypes can be rescued by UAS-Flippase or by introducing a genomic rescue construct (Dp(1;3)DC313 (Venken et al. 2010), Figure 1C) that covers the sl locus (Figure 2B and Figure 2C). These data show that all the observed phenotypes in slT2A mutants can be attributed to the loss of sl.

slT2A is a loss-of-function allele that affect fly wing and eye development

(A) sl expression in wing and eye dises. Expression of UAS-mcherry.nls (red) was driven by slT2Ato label the nuclei of the cells that expressed sl. sl is expressed in the 3rd instar larval wing disc (left) and eye disc (right). Higher magnification image of the wing disc pouch region indicated by dashed rectangle is shown. The posterior/anterior and dorsal/ventral compartment boundaries are indicated by dashed lines in yellow. Scale bars, 100μm

(B) slT2A causes a wing size reduction and ectopic veins (arrowhead) in hemizygous mutant male flies. The wing phenotypes can be rescued by introduction of a genomic rescue (GR) construct or the expression of Flippase. Scale bars, 0.5mm. The quantification of adult wing size is shown in the right panel. Each dot represents the measurement of one adult wing sample. Unpaired t test, ∗∗∗∗p < 0.0001, mean ± SEM.

(C) slT2A causes extra photoreceptors (arrows) in the hemizygous mutant flies. The eye phenotype can be rescued by introduction of a genomic rescue (GR) construct. The photoreceptor rhabdomeres stain positive for phalloidin labeling F-actin. Scale bars, 10μm. The quantification is shown in the right panel. Each dot represents the measurement of one retina sample. Unpaired t test, ∗∗∗∗p < 0.0001, mean ± SEM.

The sl gene is expressed in the fly CNS and loss of sl causes longevity and locomotion defects

Given that human PLCG1 is highly expressed in the central nervous system (CNS) (Consortium 2015) and that the affected individuals present with neurologic phenotypes including hearing or vision deficits (Table 1), we investigated the expression pattern and the cell type specificity of sl in the CNS of flies. sl is expressed in the larval CNS as well as the adult brain, and co-staining with the pan-neuronal marker Elav (Robinow and White 1991) and glial marker Repo (Sepp et al. 2001) show that sl is expressed in many neurons and glia cells in the CNS (Figure 3A). We therefore assessed the longevity and climbing of slT2A flies. Compared to the wild-type w1118 flies, slT2A/Y hemizygous mutant flies show a shortened lifespan and a progressively reduced climbing ability. These phenotypes can be rescued by expression of the wild-type sl cDNA(slT2A/Y; UAS-slWT) (Figure 3B).

sl is expressed in a subset of neurons and glia in the CNS, and loss of sl causes behavioral defects

(A) Expression pattern of sl in the central nervous system observed by slT2A-driven expression of UAS-mCherry.nls reporter (red). In either larval or adult brain, sl is expressed in a subset of fly neurons and glia, which were labeled by pan-neuronal marker Elav (green, upper panel) and pan-glia marker Repo (green, lower panel). Higher magnification images of the regions indicated by dashed rectangles are shown. Scale bars, 20μm in the magnified images, 50μm in other images.

(B) Loss of sl causes defects in longevity and locomotion. slT2A hemizygous flies have a shorter lifespan than w1118 control flies. The median lifespan of slT2A and w1118 flies is 40 days and 62 days respectively. The shorter lifespan of slT2A flies can be rescued by a UAS transgene that expresses the wild-type sl cDNA (slWT). Fly locomotion was assessed by climbing assay (see methods). slT2Aflies at the age of 7 days show reduced locomotion and become almost immotile at the age of 35 days. The reduced locomotion ability in slT2A flies can be fully rescued by slWT. For lifespan assay, Longrank test, ****p<0.0001. For climbing assay, each dot represents measurement of one vial containing 18-22 flies for test. Unpaired t test, ****p<0.0001.

Functional assays in flies indicate that the PLCG1 variants are toxic

To assess the impact of the variants, we expressed the sl variant cDNAs in the slT2A/Y hemizygous mutant males(slT2A/Y; UAS-slvariants) and compared their rescue ability with the wild-type sl(slT2A/Y; UAS-slWT). As shown in Figure 4A (middle panel), the slT2A/Y mutant flies (or the ones expressing a UAS-Empty control construct) have a slightly reduced eclosion rate, but expression of the slWT cDNA fully rescues the percentage of eclosing progeny as measured by the Mendelian ratio. In contrast, expression of slL630F (slT2A/Y; UAS-slL630F) reduced the percentage of hemizygous male progeny from the expected 25% to approximately 17%, while expression of slH384R causes a severe reduction in the number of eclosing flies, with only a few escapers (slT2A/Y; UAS-slH384R). Expression of the slD1041G or slD1184G leads to 100% lethality. These data clearly indicate that these variants are toxic but at different levels.

The human and corresponding fly variants are toxic when expressed in flies

(A) Summary of the viability associated with expression of sl cDNAs in slT2A mutant or heterozygous flies. Cross strategy: heterozygous slT2A female flies were crossed to male flies carrying UAS-cDNAs or control (UAS-Empty) constructs, or crossed to the y w males as an extra control. The percentages of hemizygous slT2A/Y male progeny (red) or slT2A/yw heterozygous female progeny (blue) that express different UAS-cDNA constructs were calculated. The expected Mendelian ratio is 0.25 (indicated by the green line in the graph). The fly analogue variants of the proband-associated variants were tested. Each dot represents one independent replicate. Unpaired t test, *p<0.05, **p < 0.01, ****p < 0.0001, ns: not significant, mean ± SEM.

(B) Summary of the viability associated with the expression of PLCG1 cDNAs in slT2A mutant (red, males) or heterozygous (blue, females) flies. The same cross strategy and progeny ratio measurement described in (A) were applied. The proband-associated variants as well as three previously reported PLCG1 variants were assessed. We also included the PLCG2 reference cDNA. Each dot represents one independent replicate. Unpaired t test, *p<0.05, **p < 0.01, ns: not significant, mean ± SEM.

Since the slT2A/Y; UAS-cDNA hemizygous males lack the endogenous sl+, we tested slT2A/yw; UAS-cDNA heterozygous female flies that carry a copy of wild-type sl+ while simultaneously expressing UAS-cDNAs driven by the slT2A driver in the cells that endogenously express sl (Figure 4A, right panel). The eclosion rates of heterozygous female progeny expressing sl variants were significantly reduced compared to those expressing slWT. Expression of slH384R or slL630F in the heterozygous progeny reduced the expected 25% proportion to approximately 10% and 20%, respectively, whereas expression of slD1041G or slD1184Gresulted in complete lethality in heterozygous flies. These results suggest that the missense variants exert a dominant toxic effect. Additionally, we observed that the toxicity may have both developmental and acute effects in adults, with varying severity among the different variants (Figure S2), indicating that sl function is required in adult flies, implying that PLCG1 variants may cause long-term deficits in affected individuals.

To compare the sl and PLCG1 associated phenotypes, we conducted similar assays using human PLCG1 cDNAs (Figure 4B). Expression of PLCG1Reference in the slT2A/Y mutant flies (slT2A/Y; UAS-PLCG1Reference) reduces viability by 80%, and expression of the other PLCγ coding gene, PLCG2, is also toxic and causes similar viability reduction compared to PLCG1Reference. This suggests that expression of human PLCγ genes is toxic in flies. This toxicity appears to be associated with expression level (Figure S3), and the survivals of slT2A/Y; UAS-PLCG1Reference did not show rescue of the loss-of-function phenotypes in the wings or eyes (Figure S4). Expression of PLCG1H380Ror PLCG1L597F in the slT2A/Y mutant flies (slT2A/Y; UAS-PLCG1H380Ror slT2A/Y; UAS-PLCG1L597F) leads to a significant but very modest increase in lethality when compared to PLCG1Reference, whereas expression of PLCG1D1019G or PLCG1D1165Gresults in 100% lethality (Figure 4B, left panel). When the reference and variants are assayed in the presence of a wild type copy of sl+, the heterozygous female progeny expressing the reference cDNA of PLCG1 or PLCG2 exhibited normal eclosion rate, as did the ones expressing PLCG1H380R or PLCG1L597F, suggesting that the presence of a wild-type copy of sl+ combined with the reduced expression levels (typically 50% due to dosage compensation for the cells on X chromosome) mask some of the potential toxicity. However, expression of PLCG1D1019G or PLCG1D1165Gstill resulted in complete lethality in the females (Figure 4B, right panel). In summary, expression of the PLCG1 variants and the corresponding fly sl variants exhibit greater toxicity than the reference or wild-type proteins with varying degrees of severity, suggesting that the variants are likely to be gain-of-function or neomorphic alleles. Among them, the PLCG1D1019G and slD1041G, as well as PLCG1D1165G and slD1184G are very strong toxic alleles whereas PLCG1H380R, PLCG1L597F, and their fly analogues are mild variants.

The p.Asp1019Gly and p.Asp1165Gly variants are hyperactive

To assess whether the variants act as gain-of-function alleles that enhance the enzymatic activity of the PLCγ1 isozyme, we tested them using a Ca²D reporter assay. Since one of the products of the PLCγ1 isozyme, IPD, binds to receptors on the endoplasmic reticulum to trigger Ca²D release (Foskett et al. 2007), intracellular Ca²D levels can serve as a proxy of the PLCγ1 enzymatic activity. We expressed the CaLexA (calcium-dependent nuclear import of LexA) reporter (Masuyama et al. 2012) in the wing disc pouch region using a specific GAL4 driver (nub-GAL4 > UAS-CaLexA.GFP) while simultaneously expressing UAS-PLCG1 cDNAs. We first assessed three control variants: PLCG1H380A, PLCG1D1165H, and PLCG1S1021F. Substitution of His380 with Ala (H380A) has been reported to suppress PIP2 hydrolysis and IP3 production (Smith et al. 1994; Wada et al. 2022), acting as an enzymatic-dead loss-of-function allele. On the other hand, the p.Asp1165His (D1165H) variant was previously identified as a strong gain-of-function somatic variant in adult T cell leukemia/lymphoma (Kataoka et al. 2015; Hajicek et al. 2019; Siraliev-Perez et al. 2022), and has been documented to cause a dramatic increase in phospholipase activity in vitro (Hajicek et al. 2019; Siraliev-Perez et al. 2022). The p.Ser1021Phe variant was reported recently in a de novo case and was characterized as a gain-of-function germline variant (Tao et al. 2023). As shown in Figure S5A, the GFP signal of the CaLexA.GFP reporter was low in wing discs expressing PLCG1H380A, whereas the signal was significantly enhanced in those expressing PLCG1D1165H or PLCG1S1021F, showing that this is a robust assay for detecting increased enzymatic activity. We next tested the variants of the affected individuals. As shown in Figure 5A, expression of PLCG1Reference did not induce obvious GFP signals, suggesting that the protein is not enzymatically active, possibly because of autoinhibition. Similarly, expression of PLCG1H380R or PLCG1L597F did not significantly alter the GFP signal, suggesting that they are not constitutively active. However, expression of PLCG1D1019G or PLCG1D1165G markedly increased the GFP signal, similar to the PLCG1D1165Hand PLCG1S1021F positive controls (Figure 5A and S4A). The same observations were made with the fly sl variants (Figure S5B). These results indicate that the PLCG1D1019G and PLCG1D1165G variants are hyperactive, whereas the PLCG1H380R and PLCG1L597F variants are not hyperactive based on this assay.

Ectopic expression of PLCG1 variants causes variable phenotypes

(A) The Ca2+ reporter CaLexA.GFP was expressed in the wing disc pouch, simultaneously with the PLCG1 cDNAs. Expression of PLCG1D1019G or PLCG1D1165Gcaused elevated CaLexA.GFP signal (green), indicating increased intracellular Ca2+ levels indicating that these variants are hyperactive. Nuclei were labeled with DAPI (blue). Scale bars, 100μm.

(B) Representative images of the adult wing blades showing the morphological phenotypes caused by wing-specific expression of PLCG1 cDNAs. Expression of PLCG1D1019G or PLCG1D1165G caused severe wing morphology defects including notched margin (arrows) and fused/thickened veins (arrowheads). Expression of PLCG1L597F exhibited partial penetrance (penetration ratio indicated). Expression of PLCG1H380R exhibited very mild phenotypes, comparable to PLCG1Reference. Scale bars, 0.5mm.

(C) Representative images showing that eye-specific expression of PLCG1Reference or PLCG1H380R causes a ∼15% eye size reduction compared to the UAS-Empty control construct, and expression of PLCG1L597F further reduced eye size. Expression of PLCG1D1019G or PLCG1D1165Gcauses a severe size reduction by ∼30%. Each dot in the quantification graph represents the measurement of one adult eye. Unpaired t test, *p<0.05, ****p < 0.0001, ns: not significant, mean ± SEM. Scale bars, 100μm.

The PLCG1 variants affect size and morphology of wings and eyes

To further assess the impact of the PLCG1 variants on normal development, we analyzed the morphology of the adult wings upon wing-specific expression of PLCG1 or sl cDNAs (nub-GAL4 > UAS-cDNAs). Interestingly, ectopic expression of either PLCG1Reference or slWT in the wing disc leads to a ∼10% reduction in adult wing size when compared to the UAS-Empty control (Figure S6A). This observation, together with the reduced wing size seen in the loss-of-function context (Figures 2B), suggests that both reduced and elevated levels of PLCγ1 can impair wing growth. This implies a dosage-dependent regulation on wing growth by the PLCγ1 isozymes, while the underlying mechanism is unknown. Additionally, as shown in Figure 5B and S6B, approximately 10% of the wings expressing PLCG1Reference exhibit notching along the wing margin, a phenotype not observed in wings expressing slWT. Expression of PLCG1H380R or PLCG1L597F caused notched wings in approximately 18% and 23% of the flies, respectively (Figure S6B), whereas expression of PLCG1D1019G or PLCG1D1165G results in severe wing phenotypes characterized by notched wing margins, fused/thickened veins and reduced wing sizes with >95% penetrance (Figure 5B). Notably, expression of fly slvariantscould lead to similar morphological defects as their corresponding human variants, arguing that these wing phenotypes are due to alterations of PLCG1 or Sl protein function (Figure 5B and S6C).

We also assessed the effect of expression of human PLCG1 on eye development using the eyeless-GAL4 (ey-GAL4). Expression of PLCG1Referenceor PLCG1H380R in fly eyes leads to a mild reduction in eye size when compared to UAS-Empty control (Figure 5C and S6D). However, expression of PLCG1L597F results in rough eyes that are reduced in size whereas overexpression of PLCG1D1019G or PLCG1D1165Gleads to a more severe eye phenotype (Figure 5C and S6D). In summary, the eye data are consistent with the wing data, showing that PLCG1D1019G and PLCG1D1165G are more toxic than PLCG1Reference. On the other hand, the toxicity of PLCG1H380R and PLCG1L597Fare stronger than the PLCG1Reference but not as severe as PLCG1D1019G and PLCG1D1165G. Interestingly, the morphological defects in wings or eyes caused by ectopic expression of PLCG1 cDNAs correlate with the expression level (Figure S6E), but do not directly correlate with the phospholipase enzymatic activity. For example, expression of PLCG1S1021F does not cause obvious morphological defects when compared to PLCG1Reference (Figure S6B and S6F), even though PLCG1S1021Fis hyperactive and induces significantly elevated intracellular Ca2+ in the CaLexA reporter assay (Figure S5A).

Discussion

Here we report seven individuals who carry heterozygous missense variants in PLCG1 which encodes the phospholipase C γ1 isozyme. The individuals present with partially overlapping clinical features including hearing impairment, eye abnormality, heart defects and immune phenotypes. We show that the fly ortholog, small wing (sl), is widely expressed in wings and eyes, as well as in the central nervous system. Consistent with its expression pattern, we report that sl not only regulates wing and eye development, as previously documented, but also plays critical roles in the nervous system and affects locomotion and longevity. Furthermore, we assessed the function of the variants in the context of the human and fly cDNAs and show that their expression induces variable levels of toxicity when compared to the reference PLCG1 or wild-type sl. Two of the variants are clearly hyperactive, and all the variants exhibit neomorphic effects (discussed in Supplemental Figure Notes). These observations show that the variants impair the normal function in vivo and suggest that they contribute to the symptoms observed in the affected individuals. Similarly to inborn error caused by the paralogous PLCG2 (Baysac et al. 2024), germline variants in PLCG1 can be pathogenic and dominant by different mechanisms.

Structural analysis of the PLCG1 variants

Previously, studies based on biochemical assays and protein structures provided insights into how the variants studied here may affect the enzymatic activity of PLCγ1 (the protein structure of full-length rat Plcg1 is shown in Figure 6A). In its basal state, the PLCγ-specific regulatory array (sPH-nSH2-cSH2-SH3) forms autoinhibitory interfaces with the catalytic domains. Upon activation by the RTKs through binding with nSH2, PLCγ1 is phosphorylated, which induces the dissociation of the inhibitory cSH2 domain from the C2 domain. This triggers conformational rearrangements, allowing the enzyme to associate with the membrane and to expose the catalytic domains to allow hydrolysis of PIP2 (Gresset et al. 2010; Hajicek et al. 2019; Liu et al. 2020; Le Huray et al. 2022; Nosbisch et al. 2022). As shown in Figure 6A, the proband-associated variants map to conserved domains of the protein, either within the catalytic domains or at intramolecular and intermolecular interfaces. The p.Asp1019Gly and p.Asp1165Gly variants impact key residues involved in autoinhibition, leading to increased enzymatic activity. Specifically, the p.Asp1019Gly variant affects a conserved residue within the hydrophobic ridge of the Y box (Figure 6B), which is important for interaction with the sPH domain. This interaction is critical for the autoinhibition by blocking the membrane engagement of the catalytic core domain prior to enzymatic activation (Ellis et al. 1998; Hajicek et al. 2019). Notably, a substitution at the same position (Asp1019Lys, D1019K) has been demonstrated to enhance basal phospholipase activity in vitro (Hajicek et al. 2013), supporting its regulatory importance. Similarly, another hotspot somatic variant, p.Ser345Phe, located in the corresponding hydrophobic ridge within the X box, is also hyperactive (Vaque et al. 2014; Manso et al. 2015). On the other hand, the p.Asp1165Gly variant affects a residue situated within a loop of the C2 domain (Figure 6C). The Asp1165 residue plays a key role in stabilizing the interaction between the cSH2 domain and the C2 domain to maintain the autoinhibited state (DeBell et al. 2007). As mentioned above, the somatic variant p.Asp1165His leads to significantly elevated phospholipase activity in vitro (Liu et al. 2020; Siraliev-Perez et al. 2022), and results in severe phenotypes in vivo (Figure 4B, S5E and S5F). Molecular dynamics simulation data consistently indicate that autoinhibition is likely disrupted by the p.Asp1019Gly and p.Asp1165Gly variants (Figure S7A). In contrast, the p.His380Arg variant impacts the His380 residue within the X box, situated near a Ca2+ ion in the catalytic core (Figure 6D). His380 plays a role in coordination of the phosphate group at the 1-position of IP3 (Le Huray et al. 2022). While this residue may not be key to autoinhibition, it is important for the phospholipase activity. Substitution of His380 with Phe or Ala (H380F, H380A) have been reported to suppress PIP2 hydrolysis and IP3 production (Smith et al. 1994; Wada et al. 2022). Hence, substitution of the His380 with Arg in p.His380Arg variant may create a more basic environment, impacting the lipase activity. On the other hand, the p.Leu597Phe variant affects a residue within the nSH2 domain, which is part of the PLCγ-specific regulatory array (Figure 6E). The nSH2 domain mediates interactions with phosphorylated tyrosine residues on RTKs to initiate activation (Bae et al. 2009). Leu597 is located near the phosphotyrosine-binding pocket and this variant may therefore alter receptor specificity or induce novel protein interactions. Additionally, we utilized the DDMut platform (Zhou et al. 2023) to predict protein stability and folding of the variants, which are discussed in Figure S7B. In summary, our in vivo data are consistent with previous reports and in silico analyses, showing that the affected amino acids map to critical residues and strengthening the conclusion that the variants are pathogenic and likely impact the protein function through distinct mechanisms.

PLCG1 variants affect important residues

(A) 3D structure of full-length rat Plcg1 (rat Plcg1 shares 97% amino acid identity with human PLCG1). The conserved protein domains are labeled with different colors. Two major intracellular interfaces are circled by dashed lines: 1-The hydrophobic ridge between the sPH domain and the catalytic core (X-box and Y-box); and 2-The interface between the cSH2 domain and the C2 domain. The four amino acids affected by the variants are shown as bolded black and indicated by yellow balls.

(B) Enlarged views of the Asp1019 residue within the autoinhibition interface between sPH domain and the Y box. The potential interactions with nearby residues are indicated.

(C) Enlarged view of the Asp1165 residue within the autoinhibition interface between the cSH2 domain and the C2 domain. The potential interactions with nearby residues are indicated.

(D) Enlarged view of the His380 residue within the X-box catalytic domain, in proximity to the Ca2+ cofactor.

(E) Enlarged view of the Leu597 and nearby residues in the nSH2 domain. Structural analysis was performed via UCSF Chimera (Pettersen et al. 2004)

The PLCG1 variants affect protein function to varying degrees and are associated with variable clinical manifestations

To better assess the genotype-phenotype relationship of the variants, we summarize the clinical features of affected individuals in Table 1, and the phenotypic effects observed in fly assays in Table S3. The p.(Asp1019Gly) (carried by Individual 1) and p.(Asp1165Gly) (carried by Individual 3) variants and their corresponding fly variants induce severe phenotypes across all assays performed. Individuals 1 and 3 share several obvious clinical features including hearing loss and heart septal defect. In contrast, the p.(His380Arg) and p.(Leu597Phe) variants cause mild or partially penetrant phenotypes across the different fly assays. Individual 2 who carries the p.(His380Arg) variant does not exhibit hearing impairment or heart defects observed in Individuals 1 and 3, but has eye malformations and neuroinflammation features that are shared with individuals 1 and 3, although the ocular and immunological defects manifest differently among individuals. Interestingly, individuals 4-7 are from the same family and all carry the p.(Leu597Phe) variant but also differ in their phenotypes, yet all share some clinical features with Individuals 1-3 (Table 1).

The heterogeneity in clinical manifestations may be influenced by additional genetic variants (see Table 1 legend) and environmental factors. Additionally, the variable expressivity observed in carriers of the same variant may be explained by allelic expression bias through autosomal random monoallelic expression (aRME) (Reinius and Sandberg 2015), a phenomenon that is thought to be common among carriers of genetic defects associated with inborn errors of immunity (IEIs). Indeed, these conditions often exhibit non-Mendelian segregation patterns and variable clinical features (Stewart et al., 2025). Moreover, the PLCγ1 isozyme is an integral component of multiple signaling pathways, and the consequences of its dysregulation are likely to be context dependent. It is likely that different PLCG1 variants impact distinct cellular processes across various tissues and cell types, resulting in a spectrum of pathological changes. In summary, the symptoms observed in affected individuals appear to correlate, to some extent, with the severity of the variants as indicated by fly assays. However, the penetrance and expressivity of these phenotypes will require further investigation to better understand the genotype-phenotype associations of PLCG1 variants.